---
title: "VCMoE Tutorial: Gaussian Inference and Count Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{VCMoE Tutorial: Gaussian Inference and Count Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE
)
```

This tutorial gives a compact workflow for using `VCMoE`. The Gaussian example
is the main walk-through because it shows fitting, diagnostics, coefficient
plots, analytic simultaneous confidence bands, and bootstrap inference. The
Binomial and Negative-Binomial examples use the same model interface and focus
on response format, prediction scale, and diagnostics.

## Installation from GitHub

Install the package from GitHub with `remotes`:

```{r install, eval=FALSE}
install.packages("remotes")
remotes::install_github("qc-zhao/VCMoE")
```

Then load the package:

```{r packages}
library(VCMoE)
library(ggplot2)
```

## Gaussian model

Simulate a small Gaussian data set with two latent components. The returned
object contains the observed data and the true coefficient functions.

```{r gaussian-simulate}
set.seed(1)

sim <- simulate_vcmoe_gaussian(
  n = 180,
  k = 2,
  seed = 1,
  separation = 1.6,
  scenario = "well_separated"
)

head(sim$data)
```

The model formula has two parts:

- `y ~ z1` is the component-specific expert mean model;
- `| x1` is the gating model for component probabilities.

The varying coordinate is supplied through `u = "u"`.

```{r gaussian-fit}
fit <- vcmoe_fit(
  y ~ z1 | x1,
  data = sim$data,
  u = "u",
  family = "gaussian",
  k = 2,
  bandwidth = 0.35,
  u_grid = seq(0.15, 0.85, length.out = 4),
  control = list(maxit = 60, n_starts = 1, seed = 2, warn_ambiguous = FALSE)
)

fit
```

Expert coefficients are returned as an array indexed by grid point, component,
and term.

```{r gaussian-coefficients}
expert_coef <- coef(fit, "expert")
dim(expert_coef)
expert_coef[, , "z1"]
```

Predictions can be requested as marginal means, posterior component
probabilities, or component-specific means.

```{r gaussian-predictions}
head(predict(fit, type = "mean"))
head(predict(fit, type = "posterior"))
head(predict(fit, type = "component"))
```

## Diagnostics and basic plots

Always inspect diagnostics before interpreting coefficient functions.

```{r gaussian-diagnostics}
diagnostics <- vcmoe_diagnostics(fit)
diagnostics[, c("u", "converged", "ambiguous", "posterior_entropy", "effective_n")]
```

`plot_coefficients()` and `plot_posterior()` provide quick visual checks.

```{r gaussian-coefficient-plot}
plot_coefficients(fit, "expert")
```

```{r gaussian-posterior-plot}
plot_posterior(fit)
```

## Analytic simultaneous confidence bands

`vcmoe_confband()` computes diagnostic-gated analytic-style Epanechnikov path
bands. The returned object contains an interval table and grid-level
diagnostics. Rows with `status != "ok"` are blocked because the local fit is too
weak for the interval to be interpreted.

```{r gaussian-scb}
band <- vcmoe_confband(
  fit,
  data = sim$data,
  level = 0.95,
  type = "simultaneous",
  coefficient_set = "expert",
  strict = FALSE
)

band
head(band$intervals[, c(
  "u", "component", "term", "block", "estimate",
  "lower", "upper", "status", "block_reason"
)])
```

For coefficient-function plots, use the local-linear intercept rows. The slope
rows describe local derivative terms and are not the coefficient functions
themselves.

```{r gaussian-scb-plot}
scb_plot <- subset(
  band$intervals,
  coefficient_set == "expert" & block == "intercept" & status == "ok"
)

ggplot(scb_plot, aes(x = u, y = estimate, color = component, fill = component)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.18, color = NA) +
  geom_line(linewidth = 0.8) +
  facet_wrap(~ term, scales = "free_y") +
  labs(
    x = "u",
    y = "coefficient",
    color = "component",
    fill = "component"
  ) +
  theme_minimal(base_size = 12)
```

## Bootstrap inference

Parametric bootstrap inference is also available. Each bootstrap replicate
simulates a new response from the fitted mixture, refits the same VCMoE model,
and aligns bootstrap component labels back to the reference fit.

```{r gaussian-bootstrap}
boot <- vcmoe_bootstrap(
  fit,
  data = sim$data,
  B = 6,
  seed = 5,
  min_successful = 2,
  control = list(maxit = 40, n_starts = 1, warn_ambiguous = FALSE)
)

boot
head(confint(boot, parm = "expert", type = "pointwise"))
```

`plot_inference()` visualizes bootstrap intervals directly.

```{r gaussian-bootstrap-plot}
plot_inference(
  boot,
  coefficient_set = "expert",
  type = "pointwise",
  level = 0.95
)
```

## Optional bandwidth selection

For real analyses, bandwidth should usually be selected rather than fixed by
hand. The selector uses held-out predictive log-likelihood and returns a final
refit by default.

```{r bandwidth, eval=FALSE}
selection <- vcmoe_select_bandwidth(
  y ~ z1 | x1,
  data = sim$data,
  u = "u",
  family = "gaussian",
  k = 2,
  bandwidth_grid = c(0.25, 0.35, 0.45),
  folds = 3,
  u_grid = seq(0.15, 0.85, length.out = 4),
  control = list(maxit = 60, n_starts = 1, seed = 3),
  seed = 4
)

selection
selection$best_bandwidth
```

## Binomial responses

For Bernoulli responses, the response column must be 0/1. Expert coefficients
are on the logit success-probability scale. Predictions with `type = "mean"`
return marginal success probabilities.

```{r binomial-bernoulli}
binom <- simulate_vcmoe_binomial(
  n = 140,
  k = 2,
  seed = 11,
  trials = 1,
  separation = 1.6
)

binom_fit <- vcmoe_fit(
  y ~ z1 | x1,
  data = binom$data,
  u = "u",
  family = "binomial",
  k = 2,
  bandwidth = 0.40,
  u_grid = seq(0.15, 0.85, length.out = 4),
  control = list(maxit = 50, n_starts = 1, seed = 12, warn_ambiguous = FALSE)
)

head(predict(binom_fit, type = "mean"))
head(predict(binom_fit, type = "posterior"))
vcmoe_diagnostics(binom_fit)[, c("u", "converged", "ambiguous", "effective_n")]
```

Grouped Binomial data use the standard R two-column response form,
`cbind(success, failure)`.

```{r binomial-grouped}
grouped <- simulate_vcmoe_binomial(
  n = 140,
  k = 2,
  seed = 13,
  trials = 8,
  separation = 1.6
)

grouped_fit <- vcmoe_fit(
  cbind(success, failure) ~ z1 | x1,
  data = grouped$data,
  u = "u",
  family = "binomial",
  k = 2,
  bandwidth = 0.40,
  u_grid = seq(0.15, 0.85, length.out = 4),
  control = list(maxit = 50, n_starts = 1, seed = 14, warn_ambiguous = FALSE)
)

coef(grouped_fit, "expert")[, , "z1"]
head(predict(grouped_fit, type = "mean"))
```

## Negative-Binomial count responses

For count data, use `family = "negative-binomial"`. Expert coefficients are on
the log mean count scale. Library size or size factors should enter through an
expert-side offset, for example `offset(log_size_factor)`.

```{r negbin}
nb <- simulate_vcmoe_negbin(
  n = 140,
  k = 2,
  seed = 21,
  separation = 1.6
)

nb_fit <- vcmoe_fit(
  y ~ z1 + offset(log_size_factor) | x1,
  data = nb$data,
  u = "u",
  family = "negative-binomial",
  k = 2,
  bandwidth = 0.45,
  u_grid = seq(0.15, 0.85, length.out = 4),
  control = list(maxit = 50, n_starts = 1, seed = 22, warn_ambiguous = FALSE)
)

coef(nb_fit, "expert")[, , "z1"]
coef(nb_fit, "theta")
head(predict(nb_fit, type = "mean"))
vcmoe_diagnostics(nb_fit)[, c("u", "converged", "ambiguous", "effective_n")]
```

The same `vcmoe_confband()` interface can be used for Binomial and
Negative-Binomial fits. This introductory tutorial only visualizes confidence
bands for the Gaussian example, where the interpretation is simplest.
